Model complexity and model choice for animal movement models

### 2016-06-08 guelph_talk/guelph_panther.Rnw
still working on guelph talk

##options(tikzMetricPackages = c("\\usepackage[utf8]{inputenc}",
##    "\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}",
##    "\\usepackage{amssymb}"))
library(ggplot2)
theme_set(theme_bw())
library("ggplot2"); theme_set(theme_classic())
library("MASS")
## use Dark2 rather than Set1 because colour #6 of Set1 is yellow (ugh/too light)
scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
library("gridExtra")
library("plyr")  ## for ldply()
library("dplyr") ## for full_join()
library("reshape2")  ## for melt()
library("GGally")
library("ggstance")
library(scales) ## for squish()
library(gridExtra) ## for grid.arrange()
library(mgcv)
library(plyr)   ## for mutate()
library(psyphy) ## for restricted-link models
library(reshape2)
library(splines)  ## for vir asymptote profile
library(deSolve)
library(Hmisc)
library(gsubfn)
@

@@

\begin{itemize}
\item observations:  \\
e.g. mass mark-recapture, longitudinal density,
direct observation, telemetry (VHF, GPS)
\item most methods provide a sequence of times and locations for each individual
\item summaries:
\begin{itemize}
\item home range \\
(convex hull, kernel density estimate, etc.)
\item root-mean-squared displacement
\item step length and turning angle
\end{itemize}
\item covariates: \\
e.g. habitat map,  \\
individual characteristics (sex, age, weight \ldots)
\end{itemize}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item Florida panther: \emph{Felix ...}
    \item severely reduced habitat, endangered
    \item small, isolated population
    \item currently recovering
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    (cute panther pictures here)
  \end{column}
\end{columns}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item how does panther movement vary by sex and life history stage \\
      (juvenile, adult, mom with kittens \ldots)
    \item how does movement mediate exposure to threats \\
      (intraspecific aggression, roadkill) ?
    \item can we predict the effects of future changes in 
      population density / population structure / habitat?
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    (more cute pictures here ... or roadkill?)
  \end{column}
\end{columns}

@@

\fixme{notation, underbraces}

\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

@@

\section[Diurnal model]{Incorporating diurnal variation}
\subsection{}

@@

\begin{frame}
\frametitle{Comparison}

Figure 1 from paper
\end{frame}

\begin{frame}
\frametitle{Goodness of fit}

Figure 2 from paper
\end{frame}

\begin{frame}
\frametitle{Temporal patterns}
Figures 3 (and 4?) from paper
\end{frame}

@@

>>>>

##options(tikzMetricPackages = c("\\usepackage[utf8]{inputenc}",
##    "\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}",
##    "\\usepackage{amssymb}"))
library("ggplot2"); theme_set(theme_bw())
library(RColorBrewer)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))

library("MASS")
## use Dark2 rather than Set1 because colour #6 of Set1 is yellow (ugh/too light)
scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
library(gridExtra)
library(plyr)  ## for ldply()
library(dplyr) ## for full_join()
library(reshape2)  ## for melt()
library(tidyr)
library(GGally)
library(ggstance)
library(scales) ## for squish()
library(gridExtra) ## for grid.arrange()
library(mgcv)
library(plyr)   ## for mutate()
library(psyphy) ## for restricted-link models
library(reshape2)
library(splines)  ## for vir asymptote profile
library(deSolve)
library(Hmisc)
library(gsubfn)
@

<<cat1sumdat,echo=FALSE,cache=TRUE>>=
L <- load("../cat1sumdat.RData")
ss <- sumdat %>% filter(grepl("HMM",type)) %>%
    mutate(deltaBIC=deltaBIC-min(deltaBIC),
           model=factor(model,
              levels=c("HMM","HMM + THsin","HMM + THquad",
                       "HMM + THblock", "HMM + THhourly"),
              labels=c("homogeneous","sinusoid","quadratic",
                       "block","hourly")))
@

@@

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item observations:  \\
e.g. mass mark-recapture, longitudinal density,
direct observation, telemetry (VHF, GPS)
\item most methods provide a sequence of times and locations for each individual
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
(sequence pix)
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item summaries:
\begin{itemize}
\item home range \\
(convex hull, kernel density estimate, etc.)
\item root-mean-squared displacement
\item step length and turning angle
\end{itemize}
\item covariates: \\
e.g. habitat map,  \\
individual characteristics (sex, age, weight \ldots)
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
(step length/turning angle pix)
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item Florida panther: \emph{Puma concolor coryi}
    \item endangered subspecies
    \item severely reduced habitat
    \item small, isolated population
    \item currently recovering
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    \includegraphics[width=0.8\textwidth]{pix/map1.jpg} \\
      {\tiny \href{http://www.peer.org/campaigns/wildlife-protection/florida-panther/ecological-needs.html}{www.peer.org}} \\
     \includegraphics[width=0.9\textwidth]{pix/kitten1.jpg}
  \end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item movement variation by sex and life history stage \\
      (juvenile, adult, mom with kittens \ldots)
    \item effects of movement on threats \\
      (intraspecific aggression, roadkill) ?
    \item predicting the effects of future changes in 
      population density / population structure / habitat
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    \includegraphics[width=0.8\textwidth]{pix/momkitten1.jpg} \\
    \includegraphics[width=0.7\textwidth]{pix/roadkill.jpg}
  \end{column}
\end{columns}

@@

\textbf{state:}
\begin{equation*}
\begin{split}
  S_t & \sim \textrm{Multinomial}(S_{t-1},\mu_{S,t}) \\
\mu_{S,t} & = \textrm{multi-logistic}(\mathbf{X}_{S,t} \boldsymbol \beta_{S} ) 
\end{split}
\end{equation*}

\textbf{emission:}
\begin{equation*}
\begin{split}
\mathbf{Z}_t & \sim \left\{ \textrm{Dist}_1(\mu_{Z_1,S_t}),
\dots, \textrm{Dist}_n(\mu_{Z_n,S_t}) \right\}
 \\
\mu_{Z_i,S_t} & = g^{-1}\left(\mathbf{X}_{Z_i,t} \boldsymbol \beta_{Z_i,S_t} \right)
\end{split}
\end{equation*}

@@

\begin{frame}
\frametitle{Model complexity (Manx shearwaters, \citex{dean_behavioural_2013})}

<<shearwater>>=
sheardat <- read.table("shearwater_loglik.dat")
names(sheardat) <- c("n","loglik")
npar <- function(n) {
    ## 2 emissions parameters per state;
    ## n*(n-1) transition parameters
    2*n+n*(n-1)
}
## don't know how many total data points;
## ~130 individuals tagged; 10-minute intervals for 2 days = 250 points
##
nobs <- 30000
sheardat2 <- mutate(sheardat,
       npar=npar(n),
       dloglik=loglik-min(loglik),
       AIC=2*npar-2*loglik,
       dAIC=AIC-min(AIC),
       BIC=log(nobs)*npar-2*loglik,
       dBIC=BIC-min(BIC)) %>%
    select(-c(AIC,BIC)) %>%
    mutate(type=ifelse(dBIC==0,"min BIC",
                ifelse(dAIC==0,"min AIC","")),
           type=factor(type,levels=c("","min AIC","min BIC")))

ggplot(sheardat2,aes(n,dloglik))+
    geom_line(colour="gray")+
    geom_point(size=5,aes(colour=type))+
    xlab("number of latent states")+
    ylab(expression(Delta~"negative log-likelihood"))+
    scale_colour_manual(name="",values=c("gray","blue","red"))
## min BIC ~ 7; min AIC ~ 10
@ 
\end{frame}

\section[Diurnal model]{Incorporating diurnal variation \citepx{li_incorporating_2015}}
\subsection{}

@@

\begin{frame}
\frametitle{Temporal patterns (step length)}

<<timeplot,out.width="0.9\\textwidth">>=
simdf <- simdat %>% dplyr::select(-obs) %>%
    group_by(time) %>% summarise_each(funs(mean))
obsdf <- simdat %>% dplyr::select(c(obs,time)) %>%
    filter(!is.na(obs)) %>% group_by(time) %>%
    summarise_each(funs(mean))
combdf <- rbind(gather(simdf,model,steplen,-time),
          data_frame(time=obsdf$time,model="obs",steplen=obsdf$obs))
mnames <- c("hmm6","hmmblock4","hmmquad5",
            "hmmsin5","hmmhourly3","obs")
mnames2 <- c("homogeneous (n=6)","block (n=4)",
             "quadratic (n=5)","sin (n=5)","hourly (n=3)","observed")
combdf2 <- combdf %>%
    filter(model %in% mnames) %>%
    mutate(model=factor(model,levels=mnames,labels=mnames2))
colvec <- c(brewer.pal(5,"Dark2"),"black")
lwdvec <- rep(c(1,4),c(5,1))
alphavec <- rep(c(1,0.2),c(5,1))
ggplot(combdf2,aes(x=time,y=steplen,colour=model, lwd=model, alpha=model)) + 
    geom_line()+ xlab("Time of day") +
    ylab(expression("Mean step length"~(log[10]*m)))+
    scale_colour_manual(values=colvec)+
    scale_alpha_manual(values=alphavec)+
    scale_size_manual(values=lwdvec)
@

\begin{frame}
\frametitle{Temporal patterns (autocorrelation)}
<<acfplot,cache=TRUE,out.width="0.9\\textwidth">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
temp <- simdat %>% dplyr::select(-time)%>%ldply(.,acffun,.id="model") %>%
    filter(model %in% mnames) %>%
    mutate(model=factor(model,levels=mnames,labels=mnames2))

ggplot(temp,
       aes(x=lag,y=ACF,colour=model, lwd=model, alpha=model))+ 
    geom_line()+ xlab("Time of day") +
    ylab("Autocorrelation") +
    scale_colour_manual(values=colvec)+
    scale_alpha_manual(values=alphavec)+
    scale_size_manual(values=lwdvec)+
    geom_hline(yintercept=0,lty=2)
@

\begin{frame}
\frametitle{Goodness of fit/model complexity}
<<adj_BIC,out.width="0.9\\textwidth">>=
(adj_BIC_plot <- ggplot(ss, aes(x=nstates,y=deltaBIC,colour=model))+
    scale_size_continuous( name="# of parameters") +
    scale_x_continuous(expand=c(0,0.5),breaks=2:7)+
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="number of latent states")+
    geom_line() + zmargin)
@

@@

\begin{frame}
\frametitle{Animal movement: open challenges}
\begin{itemize}
\item Cognition
\item Intraspecific interaction/collective movement
\item Continuous-time movement models
\item Edges, barriers, and corridors
\item Efficient (big-data) approaches
\end{itemize}
\end{frame}
